On-line Vision-based Engagement Assessment for Interactive Robots

### 2015-08-28 paper.tex
First paper draft

While new algorithms and robotic technologies ensure better cognitive and interactive abilities, a deeper study about the benefits in the social interaction needs to be done: Is not enough to exhibit human-like capabilities in social tasks, but to do it in a human-like manner. For instance, several studies showed that humans prefer to interact with a robots whose task outcome is delayed, shows uncertainty or an undesired output \cite{Admoni,Short}. In fact, the ability to adapt has been proved to be beneficial in long-term interactions \cite{Tielman:2014, Lim:2014}, specifically with children, manifesting a more positive attitude over time.

@@

Engagement describes a cognitive, affective (specifically intrinsic motivation), and behavioral state of interaction with a computer application that ``makes the user want to be there" \cite{OBrien:2010}. Engaging inter-actions were thought to involve attention, intrinsic interest, interactivity, perceived control and choice, functionality and motivation.

@@

The focus of this research is to evaluate the quality of the interaction using the VFOA as an engagement indicator and the current performed task in a complex scenario using an RGB camera. The present studies has been developed in the context of the CoWriter project \footnote{The primary repository is \url{https://github.com/chili-epfl/cowriter_letter_learning}.}. However, in a child-robot interaction context, avoiding 'invasive' technologies such as wearable devices is a must to keep the experience as natural as possible.

@@

In this research, we address the problem of tracking the children's engagement based on their VFOA and current performed task during a long term child-robot interaction in a learning context, first. And second, we provide qualitative and quantitative evaluation of the outcome provided using a first person RGB camera approach. Moreover, this monitoring activity has been supporting our research in child-robot interaction for learning and instruction.
\\\\

@@

We propose a adaptive model based solution to approach head pose estimation that finds correspondences between facial landmarks points and their respective locations in a face model. We use the open-source library dlib\footnote{http://dlib.net} \cite{dlib09}, to detect and extract the skeleton of the subject's face and then, use the obtained landmarkers to calculate the head pose orientation estimation (see figure \ref{system}). In this way, we can estimate what the subject was looking at and obtain a 3D scene representation as it is executed as a ROS (Robot Operating System) node that can be seen in figure \ref{rviz}. Compared to intrusive hardware like head mounted displays, eye-trackers, wearables for instance, this approach has a clear advantage of simplicity and integration with robotic systems, but also becomes accurate enough for general purposes in face-to-face interactions. In addition, it allows as to model complex environments, handling different number and positions of participants in the scene.

@@

\subsection{Estimating Head Pose}
The first step is to estimate the head pose using the pin-hole model shown in equation \ref{eq:pinHole}. As we mention before dlib library facilitates the task by providing the 2D points \textit{m'} that composes the face silhouette. However, it is enough with obtaining eight spread points, such as right and left eyes, right and left tragions, sellion or nasal depression zone, pronasale or center of the nose, stomion or center of the mouth and menton.

Likewise, the 3D face model is needed. The proposed head pose estimation method would work best with a 3D face model of the query subject itself. When a 3D face model of the subject is not available, as in many practical situations, a 3D face model of a subject can be used. This is an estimation of the same anthropometric points \textit{M'} (for North American Caucasians) in the 3D space \cite{farkas1994anthropometry}. Moreover, the intrinsic camera parameters are necessary to be able to compose its matrix \textit{A}. The components are expanded in equation \ref{eq:pinHole2}.

@@

Finally, the matrix of extrinsic parameters [R|t] is obtained by joining the rotation matrix R (calculated from the rotation vector) and translation vector \textit{t}. The result is a matrix projection \textbf{P} which maps a point in the 3D space onto a point in the 2D image place solving the equation \ref{eq:p}.

@@

\subsection{Defining the Field of View}
The third step is to generate a representation of the field of view which origin and orientation are the ones defined by the face detection \textit{tf} frame.

@@

Assuming that the \textit{tf} frames representing the robotic agent, the two tablets, the experimenter and the observers are static, need to be monitored. This task consists on evaluating the intersection between the \textit{tf} frames and the field of view: If one focus of attention is within the region defined by the field of view, we assume the subject is looking at it. In order to solve such operation mathematically it is necessary to compute the transformation matrix between the coordinate frames to be able to locate the monitored frames \textit{A} in the subject's coordinate system \textit{B} (see equation \ref{eq:transform}).

@@

\section{Case Study}
The presented method have been successfully employed in a real child-robot interaction situation where the robot tries to engage the child in handwriting tasks using the paradigm learning by teaching.
\subsection{Experimental Set-up}

The target of this study were 6 children with ages compressed between 5 and 6 years old. The experiments were developed individually and consisted on a writing activity by turns and one story telling performed by the robot at a given time of the interaction. All this performed in a over all 20 minutes interaction average time.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.7\columnwidth]{drawSetup}
    \caption{\small Graphical representation of he set-up and the possible children focuses of attention in the scene, in red.}
    \label{drawSetup}
\end{figure}

@@

\begin{enumerate}
\item Which head pose estimation tracking rate can we achieve during a real case scenario?  
\item How accurate can we predict where the subject is looking at based on the head pose estimation?
\item How can we evaluate the performance of the children based on the focus of attention?
\end{enumerate}

\subsection{From Face Detection to Focus of Attention}

@@

In order to calculate the inter-rater reliability between the two measurements (real-captured and real-expected), the overlap percentage between the two curves has been computed. A more reliable measurement in terms of time shifting between the real and captured focuses of attention is provided using the cosine similarity. These results are shown in table \ref{tab:results}.

@@

The results has shown that the percentage of accuracy in the tracking is not only conditioned by the subject's amount of movement, but also by the speed of the same movements: Restless children externalize greater levels of movement decreasing the tracker accuracy. Additionally, several recurrent computer vision challenges such as the lighting conditions, occlusions and the face angle with respect to the camera are factors that contribute to reduce the accuracy.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.8\columnwidth]{bitmap}
    \caption{\small Cosine similarity matrix between real and obtained VFOA using a 10 sec. window size computation. The correspondent numbers are shown in figure \ref{fig:realCaptured}.}
    \label{fig:bitmap}
\end{figure}

@@

@@

Once the turn-taking is established, a pattern becomes visible during the subject's feedback, when an attempt is performed: Such pattern shows a high intermittent gaze frequency from the tablet to the selection. It also reveals that the subject is using the demonstration model in the selection tablet to provide a better answer to the robot. In fact, subjects decrease the response time after each shift as well as the gaze frequency towards the model suggesting an assimilation over time. Moreover, when a new demonstration is selected for playing, the same behavior is replicated (see yellow boxes in figure \ref{fig:realExpected}).

If we consider the previous case, evaluating the subject's progression together with the time on task may be an indicator of engagement: If the subject answers so quickly with respect to the previous attempts with a poorer handwriting may be the case she is not engaged anymore. In such hypothetical case an activity switch can be easily implemented in a robot architecture. It confirms the findings reported in \cite{anzalone} where time response to event is proposed to account quality of interaction.

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.9\columnwidth]{heatmap}
    \caption{\small 2D histogram of a subject's head movements.}
    \label{heatmap}
\end{figure}

@@

However, a high amount of gaze shift \ref{fig:gazeR} towards the demonstration provided can reveal that the challenge is too big or no gaze, the exactly the opposite. In this case, the robot response can be adapted to the situation. Similar, but least relevant results are obtained using the correctness \ref{fig:errorR} of the child's feedback after each attempt.

@@

It is also important to notice that the initiative of the robot into the learning process in a turn-taking activity produces an effect in the engagement of the human partner. However, experience suggests that a robot that learns evoke a positive action in the partner. But the opposite, a robot that is not able to progress over time, produces the opposite effect.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.9\columnwidth]{errorR}
    \caption{\small Linear regression of all subject attempts against the word correctness, measured as the average euclidean distance of each leter in its corresponding eigenspace with respect to a reference leter. The gray region defines the confidence interval.}
    \label{fig:errorR}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.9\columnwidth]{gazeR}
    \caption{\small Linear regression of all subject attempts against the occurrences of a gaze shift from the writing tablet towards the model provided. The gray region defines the confidence interval.}
    \label{fig:gazeR}
\end{figure}

\section{Conclusions}
In this work we have presented an on-line vision-based engagement assessment solution that can be extended to work in different platforms for simulating complex interaction scenarios monitoring several VFOA with a maker-less technology. It provides some benefits like the decrement of the system complexity as well as the increment of the naturalness of the interaction.

@@

Additionally, both the gaze shift and the correctness of the feedback seem to be strictly correlated with the learning gain during the activity as they go in decrement after each attempt provided by the subjects. Despite the fact that the results are encouraging, there is a need of validation with a greater amount of subject's.

Regarding to the approach used, some limitations need to be considered. The most relevant one, the fact that the positions of the VFOA need to be known in advance and defined as static \textit{tf} frames. Moreover the gaze information acquired during the interactions can also be useful to develop actively the concept gaze-aware robot, where not only the robot knows where the subject looks at, but shares the same subject's VFOA when the activity state allows it.

Moreover, other limitations are based on the fact that the eye gaze information is neglected which induces a certain amount of error, as well as the noise induced by the 3D pose estimation computed from face landmarkers.

>>>>

While new algorithms and robotic technologies ensure better cognitive and interactive abilities, a deeper study about the benefits in the social interaction needs to be done: Is not enough to exhibit human-like capabilities in social tasks, but to do it in a human-like manner. For instance, several studies showed that humans prefer to interact with robots whose task outcome is delayed, shows uncertainty or an undesired output \cite{Admoni,Short}. In fact, the ability to adapt has been proved to be beneficial in long-term interactions \cite{Tielman:2014, Lim:2014}, specifically with children, manifesting a more positive attitude over time.

@@

Engagement describes a cognitive, affective (specifically intrinsic motivation), and behavioral state of interaction with a computer application that ``makes the user want to be there" \cite{OBrien:2010}. Engaging interactions were thought to involve attention, intrinsic interest, interactivity, perceived control and choice, functionality and motivation.

@@

The focus of this research is to evaluate the quality of the interaction using the VFOA as well as the correctness in the performed task as engagement indicators in the context of the CoWriter project \footnote{The primary repository is \url{https://github.com/chili-epfl/cowriter_letter_learning}.}. In order to do that avoiding the use of 'invasive' technologies such as wearable devices becomes a must to keep the experience as natural as possible.

@@

In this research, we address the problem of tracking the children's engagement based on their VFOA and current performed task during a long term child-robot interaction in a learning context, first. And second, we provide qualitative and quantitative evaluation of the outcome provided using a first person RGB sensor approach. Moreover, this monitoring activity has been supporting our research in child-robot interaction for learning and instruction purposes.
\vfill

@@

We propose a adaptive model based solution to approach head pose estimation that finds correspondences between facial landmarks points and their respective locations in a face model. We use the open-source library dlib\footnote{http://dlib.net} \cite{dlib09}, to detect and extract the skeleton of the subject's face and then, use the obtained landmarkers to calculate the head pose orientation estimation (see figure \ref{system}). In this way, we can estimate what the subject was looking at and obtain an \textit{rviz} 3D scene representation as it is executed as a ROS (Robot Operating System) node that can be seen in figure \ref{rviz}. Compared to intrusive hardware like head mounted displays, eye-trackers or wearables for instance, this approach has a clear advantage of simplicity and integration with robotic systems, but also becomes accurate enough for general purposes in face-to-face interactions. In addition, it allows us to model complex environments, handling different number and positions of participants in the scene.

@@

\subsection{Estimating Head Pose}
The first step is to estimate the head pose using the pin-hole model shown in equation \ref{eq:pinHole}. As mentioned above, dlib library facilitates the task by providing the 2D points \textit{m'} that composes the face silhouette. However, it is enough with obtaining eight spread points, such as right and left eyes, right and left tragions, sellion or nasal depression zone, pronasale or center of the nose, stomion or center of the mouth and menton.

Likewise, the 3D face model is needed. The proposed head pose estimation method would work best with a 3D face model of the query subject itself. When a 3D face model of the subject is not available, as in many practical situations, a generic 3D face model can be used. This is an estimation of the same anthropometric points \textit{M'} (for North American Caucasians) in the 3D space \cite{farkas1994anthropometry}. Moreover, the intrinsic camera parameters are necessary to be able to compose its matrix \textit{A}. The components are expanded in equation \ref{eq:pinHole2}.

@@

Finally, the matrix of extrinsic parameters [R|t] is obtained by joining the rotation matrix R (calculated from the rotation vector) and translation vector \textit{t}. The result is a matrix projection \textbf{P} which maps a point in the 3D space onto a point in the 2D image plane solving equation \ref{eq:p}.

@@

\subsection{Defining the Field of View}
The third step is to generate a representation of the field of view whose origin and orientation are the ones defined by the face detection \textit{tf} frame.

@@

Assuming that the \textit{tf} frames representing the robotic agent and the VFOA candidates are static, need to be monitored. This task consists on evaluating the intersection between the \textit{tf} frames and the field of view: If one focus of attention is within the region defined by the field of view, we assume the subject is looking at it. In order to solve such operation mathematically it is necessary to compute the transformation matrix between the coordinate frames to be able to locate the monitored frames \textit{A} in the subject's coordinate system \textit{B} (see equation \ref{eq:transform}).

@@

\section{Case Study}
The presented method have been successfully employed in a real child-robot interaction situation where the robot tries to engage the child in handwriting tasks using learning by teaching paradigm.
\subsection{Experimental Set-up}

The target of this study were 6 children with ages compressed between 5 and 6 years old. The experiments were developed individually and consisted on a writing activity by turns and one story telling performed by the robot at a given time of the interaction. All this performed in a over all 18 minutes interaction average time.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.7\columnwidth]{drawSetup}
    \caption{\small Graphical representation of the set-up and the possible children focuses of attention in the scene, in red.}
    \label{drawSetup}
\end{figure}

@@

\begin{enumerate}
\item Which head pose estimation tracking rate can we achieve during a real case scenario?  
\item How accurate can we predict where the subject is looking at based on the head pose estimation?
\item How can we evaluate the engagement of the children based on the focus of attention?\newline
\end{enumerate}

\subsection{From Face Detection to Focus of Attention} \label{fromFaceTo}

@@

In order to calculate the inter-rater reliability between the two measurements (real-captured and real-expected), the overlap percentage between the two curves has been computed. A more reliable measurement in terms of time shifting between the real and captured focuses of attention is provided using the cosine similarity also shown in table \ref{tab:results}.

@@

The results has proved that the VFOA prediction accuracy in the tracking is not only conditioned by the subject's amount of movement, but also by the speed of the same movements: Restless children externalize greater levels of movement decreasing the VFOA prediction accuracy.

For instance, subjects 1,4 and 5 despite the fact that the tracking accuracy is in the average, the bad precision in the captured VFOA suggests that such children are highly likely to modify the gaze direction without head move intervention reducing dramatically the overlap between captured and real VFOA.

Additionally, several recurrent computer vision challenges such as the lighting conditions, occlusions with the hands and the face angle with respect to the camera, looking to the experimenter for instance, are factors that contribute to reduce the tracking accuracy.

@@

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.8\columnwidth]{bitmap}
    \caption{\small Cosine similarity matrix between real and cap-tured VFOA using a 10 sec. window size computation. The correspondent numbers are shown in figure \ref{fig:realCaptured}.}
    \label{fig:bitmap}
\end{figure}

@@

Once the turn-taking is established, a pattern becomes visible during the subject's feedback, when an attempt is performed: Such pattern shows a high intermittent gaze frequency from the tablet to the selection. It also reveals that the subject is using the demonstration model in the selection tablet to provide a better answer to the robot. In fact, subjects decrease the response time after each shift as well as the gaze frequency towards the model suggesting an assimilation over time. Moreover, when a new demonstration is selected for playing, the same behavior is replicated (see yellow boxes in figure \ref{fig:realExpected}).
\\
\subsection{Focus of attention distribution}
Additional information of the head pose can be obtained for the evaluation of the VFOA distribution. Figure \ref{heatmap} show the most salient focuses of attention where the peaks are located. These peaks are mainly projected between the triangle composed by the robot, the writing tablet and the selection tablet using an unsupervised K-means clustering method as figure \ref{kmeans} reveals, considering the observers as negligible.

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.9\columnwidth]{heatmap}
    \caption{\small 2D histogram of subject 4 head movements. (head yaw on x-axis, pitch on y-axis). In red, peaks representing the most salient zones.}
    \label{heatmap}
\end{figure}

@@

\subsection{Engagement assessment}
If we consider the case shown in section \ref{fromFaceTo}, evaluating the subject's progression together with the time on task may be an indicator of engagement: If the subject answers so quickly with respect to the previous attempts with a poorer handwriting may be the case she is not engaged anymore. In such hypothetical case an activity switch can be easily implemented in a robot architecture. It confirms the findings reported in \cite{anzalone} where time response to event is proposed to account quality of interaction.

However, a high amount of gaze shift (figure \ref{fig:gazeR}) towards the demonstration provided can reveal that the challenge is too big or no gaze, the exactly the opposite. In this case, the robot response can be adapted to the situation. Similar, but least relevant results are obtained using the correctness (figure \ref{fig:errorR}) of the child's feedback after each attempt.

\begin{figure}
    \centering
    \includegraphics[width=0.9\columnwidth]{gazeR}
    \caption{\small Linear regression of all subject attempts against the occurrences of a gaze shift from the writing tablet towards the model provided. The gray region defines the confidence interval.}
    \label{fig:gazeR}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.9\columnwidth]{errorR}
    \caption{\small Linear regression of all subject attempts against the word correctness, measured as the average euclidean distance of each leter in its corresponding eigenspace with respect to a reference leter. The gray region defines the confidence interval.}
    \label{fig:errorR}
\end{figure}

@@

It is also important to notice that the initiative of the robot into the learning process in a turn-taking activity produces an effect in the engagement of the human partner. However, experience suggests that a robot that learns evoke a positive action in the partner. But the opposite, a robot that is not able to progress over time, produces the contrary effect.

@@

\section{Conclusions}
In this work we have presented an on-line vision-based engagement assessment solution that can be extended to work in different platforms for simulating complex interaction scenarios monitoring several VFOA with a maker-less technology. It provides some benefits like the decrement of the system complexity as well as the increment of the naturalness of the interaction due to the use of an RGB sensor.

@@

Additionally, both the gaze shift and the correctness of the feedback seem to be strictly correlated with the learning gain during the activity as they go in decrement after each attempt provided by the subjects. Despite the fact that these results are encouraging, there is a need of validation with a greater amount of subjects.

Regarding to the approach used, some limitations need to be considered. The most relevant one, the fact that the positions of the VFOA need to be known in advance and defined as static \textit{tf} frames in the system. Moreover, it is important to notice that eye gaze information is neglected which induces a certain amount of error, as well as the noise induced by the 3D pose estimation computed from face landmarkers.

Gaze information acquired during the interactions is not only useful to detect and predict engagement, but also allows the possibility to develop actively the concept gaze-aware robot. In other words, make the system able to estimate where the subject looks at, but also imitate the same subject's VFOA when the activity state allows it.
